---
title: "Visualizing data in R"
format: html
  # closeread-html:
  #   remove-header-space: true
---

## Getting started

In this page we will build on our work in the previous page, where we learned how to read-in data and conduct some basic data manipulation operations.

## Getting back to where we were

Last time, we saved our work in a file called "nhis_working_data.csv" in a directory that you chose (or the default directory). We will read in that file using the same `read_csv` command that we used to download the original data. Before we can do that, however, we need to acitvate the `tidyverse` packages:

> library(tidyverse) data\<-read_csv("nhis_working_data.csv")

```{r}
#| echo: false
library(tidyverse)
data<-read_csv("nhis_working_data.csv")

```

We can see from the output that we have attached the `tidyverse` packages and opened the NHIS file that we were working on last time. To double check, make sure that you have the variables `over65` and `age_minus_65` in `data`.

## Visualizaing data

### Plotting point-based data

We can use the `ggplot` function to plot data. We will start by plotting the relationship between age and insurance coverage. `ggplot` uses a set of ideas that can be roughly classified as data, aesthetics, and geometries. We feed in the data to ggplot, tell it which variables correspond to which aesthetic elements, and use a geometry to plot the data. For example, for "point" or "scatter" plots, we use the `geom_point` geometry which requires the aesthetics `x` and `y`. Here's the code to plot the age pattern of insurance coverage:

> data\|\>ggplot(aes(x=age, y=any_insurance))+geom_point()

In this code, `ggplot` says that we are creating a new plot; the `aes` function let's us define our aesthetics, and `geom_point` says that we are plotting point data. The `x` aesthetic is the `age` variable and the `y` aesthetic is the `any_insurance` variable. The plot looks like this:

```{r}
#| echo: false
data|>ggplot(aes(x=age, y=any_insurance))+geom_point()
```

There are a lot of points there! That is because `data` contains an observation for each age-year-sex cell; let's `summarize` our data to the age level (remember that `pop` holds the size of each cell).

> data\|\>group_by(age)\|\>summarize(any_insurance=weighted.mean(any_insurance, pop))\|\>ggplot(aes(x=age, y=any_insurance))+geom_point()

```{r}
#| echo: false
data|>group_by(age)|>summarize(any_insurance=weighted.mean(any_insurance, pop))|>ggplot(aes(x=age, y=any_insurance))+geom_point()
```

This graph is much easier to read and demonstrates that insurance coverage is higher among older individuals and children than people in the "middle" age range.

### Plotting line-based data

In order to dive into the patterns we are seeing in the insurance coverage data in more detail, we will first switch to plotting these data as lines charts rather than scatter plots. We can make this switch by replaceing `geom_point` with `geom_line`:

> data\|\>group_by(age)\|\>summarize(any_insurance=weighted.mean(any_insurance, pop))\|\>ggplot(aes(x=age, y=any_insurance))+geom_line()

```{r}
#| echo: false
data|>group_by(age)|>summarize(any_insurance=weighted.mean(any_insurance, pop))|>ggplot(aes(x=age, y=any_insurance))+geom_line()
```

This line chart shows the same basic pattern as the scatter plot. Now we will make more sophisticed use of `geom_line` to help us see what is happening in insurance coverage by age. Specifically, we do not have to specify all of the aesthetics in ggplot; we can also specify them in the `geom_line` command AND we can combine several of them together using the `+` sign. Let's plot `any_insurance`, `mcare_insurance`, and `private_insurance` first (note the use of `across` to get the weighted mean for each insurance type):

> data\|\>group_by(age)\|\>summarize(across(c(any_insurance,mcare_insurance,private_insurance), \~weighted.mean(., pop)))\|\>ggplot(aes(x=age))+geom_line(aes(y=any_insurance))

```{r}
#| echo: false
data|>group_by(age)|>summarize(across(c(any_insurance,mcare_insurance,private_insurance), ~weighted.mean(., pop)))|>ggplot(aes(x=age))+geom_line(aes(y=any_insurance))+geom_line(aes(y=mcare_insurance))+geom_line(aes(y=private_insurance))
```

Hmm... this doesn't look so easy to read because all of the lines look identical. We can fix that by telling `geom_line` that there is an aesthetic `color` that will take on various values.

> data\|\>group_by(age)\|\>summarize(across(c(any_insurance,mcare_insurance,private_insurance), \~weighted.mean(., pop)))\|\>ggplot(aes(x=age))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))

```{r}
#| echo: false
data|>group_by(age)|>summarize(across(c(any_insurance,mcare_insurance,private_insurance), ~weighted.mean(., pop)))|>ggplot(aes(x=age))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))
```

Now we can see that the increase in insurance coverage at age 65 is from Medicare coverage, which rises from essentially 0% to almost 100% after people turn 65. However, this doesn't explain what is happening among kids. Let's expand the insurance types we are considering to include CHIP and Medicaid too:

> data\|\>group_by(age)\|\>summarize(across(c(any_insurance,mcare_insurance,private_insurance), \~weighted.mean(., pop)))\|\>ggplot(aes(x=age))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))

```{r}
#| echo: false
data|>group_by(age)|>summarize(across(c(any_insurance,mcare_insurance,private_insurance,chip_insurance,mcaid_insurance), ~weighted.mean(., pop)))|>ggplot(aes(x=age))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))+geom_line(aes(y=chip_insurance,color="CHIP"))+geom_line(aes(y=mcaid_insurance,color="Medicaid"))
```

This graph demonstrates that the decline in insurance coverage around 18 years of age comes from reductions in Medicaid and, to a lesser extent, CHIP coverage.

Looking at the line plots, it appears that there are three main groups that we would like to think about: children, adults, and the elderly. Let's define a variable, `age_group` that assigns each cell to one of these three groups. We will use a new command, `case_when` to do this. `case_when` is a function that allows us to create new variables based on a series of conditions. Here is the code to create the `age_group` variable:

> data\|\>mutate(age_groups=case_when(age\<=18~"Children",age\<=64~"Adults",TRUE\~"Elderly"))

Remember to assign the results back to data!

Check your work by `count`ing these levels:

```{r}
#| echo: false
data<-data|>mutate(age_groups=case_when(age<=18~"Children",age<=64~"Adults",TRUE~"Elderly"))
data|>count(age_groups)
```

Now that we have the `age_group` variable, we can use it to plot the insurance coverage data by age group over time. We will use `year` as the `x` aesthetic and `age_group` as the `color` aesthetic.

> data\|\>group_by(age_groups,year)\|\>summarize(across(c(any_insurance,mcare_insurance,private_insurance), \~weighted.mean(., pop)))\|\>ggplot(aes(x=year, group=age_group))+geom_line(aes(y=any_insurance)

```{r}
#| echo: false
data|>group_by(age_groups,year)|>summarize(across(c(any_insurance,mcare_insurance,private_insurance), ~weighted.mean(., pop)))|>ggplot(aes(x=year, color=age_groups))+geom_line(aes(y=any_insurance))
```

This graph shows that insurance coverage has been increasing for children throughout this time period and that there was a dramatic increase in insurance coverage for adults in 2014.

### Plotting with multiple panels

We can use the `facet_wrap` function to create multiple panels in a single plot. This function takes a formula that specifies the variable that we want to use to create the panels. We will make a plot that shows each of our age groups and allows us to see how insurance coverage has changed over time for each type of coverage.

> data\|\>group_by(age_groups,year)\|\>summarize(across(c(any_insurance,mcare_insurance,private_insurance,chip_insurance,mcaid_insurance), \~weighted.mean(., pop)))\|\>ggplot(aes(x=year))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))+geom_line(aes(y=chip_insurance,color="CHIP"))+geom_line(aes(y=mcaid_insurance,color="Medicaid"))+facet_wrap(\~age_groups)

```{r}
#| echo: false
data|>group_by(age_groups,year)|>summarize(across(c(any_insurance,mcare_insurance,private_insurance,chip_insurance,mcaid_insurance), ~weighted.mean(., pop)))|>ggplot(aes(x=year))+geom_line(aes(y=any_insurance,color="Any"))+geom_line(aes(y=mcare_insurance, color="Medicare"))+geom_line(aes(y=private_insurance,color="Private"))+geom_line(aes(y=chip_insurance,color="CHIP"))+geom_line(aes(y=mcaid_insurance,color="Medicaid"))+facet_wrap(~age_groups)
```

## Practice

Here are some exercises that you can use as practice to help you make sure that you have understood the concepts. In some cases you will need to look at the helpfiles. You can do this by typing `?function_name` in the console.

1.  Create a line plot that shows the relationship between age and the proportion of people with insurance coverage before and after the ACA was implemented. Please use `facet_wrap` to create a two panel graph and show trends for each of the listed insurance types.

2.  Look at the help for `geom_bar` and `geom_col`. Choose the appropriate function (either will work, but will require different setup steps). Create a bar plot that shows the proportion of people with each insurance type by before vs. after the ACA, our three age groups, and gender.
